In-silico molecular enrichment and clearance of the human intracranial space

### 2025-01-29 paper/main.tex

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to
  human brain health and function, with alterations often linked to
  disease and functional impairment, and with distinct opportunities
  for personalized diagnostics and treatment. However, our
  understanding of these mechanisms and their interplay remains
  incomplete, in part due to the complexity of integrating insights
  across scales, between species and from different modalities. Here,
  we combine geometric model reduction and mixed-dimensional
  modelling, multi-modal magnetic resonance images, and high
  performance computing to construct and explore a high-fidelity
  in-silico model of human intracranial molecular enrichment and
  clearance. This model predicts the temporal and spatial spreading of
  a solute within an image-derived geometric representation of the
  subarachnoid space, ventricular system and brain parenchyma,
  including branching networks of surface perivascular spaces
  (PVSs). Our findings highlight the significant impact of
  cerebrospinal fluid (CSF) production and intracranial pulsatility on
  molecular enrichment following intrathecal tracer injection. We
  demonstrate that low-frequency vasomotion induces moderate CSF flow
  rates in surface PVS networks that substantially enhance tracer
  enrichment, and that impaired enrichment is a natural direct
  consequence of enlarged PVSs. This openly available technology
  platform thus provides an opportunity for integrating separate
  observations on diffusion in neuropil, vascular dynamics,
  intracranial pulsatility, CSF production, and efflux, and for
  exploring questions relating to drug delivery and clearance in the
  human brain environment.
\end{abstract}
\begin{document}

@@

Our periarterial network extends from the internal carotid arteries
and basilar artery through up to 18 bifurcations to reach upstream
network ends located within the SAS or up to 6 mm inside the
parenchyma (\Cref{fig:results1}A--C, \Cref{fig:pvs_a}A) and includes
major surface arteries of radius 0.5--1.4 mm. Imposing the pressure
field induced by CSF production at the periarterial network ends
induces ultraslow steady CSF flow of variable direction in these PVSs,
with an average velocity of 0.08 \textmu m/s (antegrade) and a maximum
velocity of 0.55 \textmu m/s (\Cref{fig:pvs_a}B). On the other hand,
traveling waves of arterial wall motion (\Cref{fig:pvs_a}C), such as the pulse wave or other vasomotion~\cite{vanveluw2020vasomotion,
  munting2023spontaneous, bojarskaite2023sleep, broggini2024long,hauglund2025norepinephrine},
also induce net directional flow in the
PVS~\cite{kedarasetti2020arterial, kedarasetti2020functional,
  coenen2021lubrication, gjerde2023directional, nozaleda2024arterial}
-- of magnitude and direction depending on the amplitude, frequency,
and length of the waves and the characteristics of the perivascular
network~\cite{gjerde2023directional}. Applying a
semi-analytic model of the net flow induced by peristalsis in
perivascular networks~\cite{gjerde2023directional} (see Methods), we
estimate that the cardiac pulse wave alone, traveling at a frequency
of 1 Hz with a wavelength of 2.0 m and a 1\% wall
displacement~\cite{jung2021novel}, will induce mainly antegrade PVS
flow with an average net velocity of 0.92 \textmu m/s while reaching up
to 7.31 \textmu m/s near larger, ventral vessels such as the MCA
(\Cref{fig:pvs_a}D). The same theory predicts that strong ultraslow
vasomotions, if traveling antegrade at 0.1Hz with a wavelength of 0.02
m and a 10\% displacement~\cite{broggini2024long}, will induce both
retrograde and antegrade net PVS flow with an average velocity of
13.05 \textmu m/s and maximum velocity 54.44 \textmu m/s
(\Cref{fig:pvs_a}E).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.99\textwidth]{figures/figure3b.png}
    \caption{\textbf{Perivascular flow shapes and accelerates molecular enrichment.}
    A) Comparison of mean tracer concentration in the PVS and over the outer PVS surface at key locations of the arterial tree, for the baseline (lower) and high PVS flow (upper) models, with labels: ``peak'': time-to-peak (h), ``$\Delta t$'': time difference between peaks (h), and ``FTA'': first-time tracer arrival (h).
    B) first-time arrival for the main trunks of the periarterial network over the distance from the closest network root node (BA, ICA-R or ICA-L for the baseline, the high PVS flow and the high PVS dispersion models);}
    \label{fig:pvs_b}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
    C, D) 2D slices showing zoom-in on the region surrounding the MCA-R and MCA2-R for the baseline (upper) and high PVS flow (lower) models. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF));  
    E) Total amount of tracer in the periarterial spaces as a function of the distance to the nearest network root over time for the baseline (left) and the high PVS flow (right) models;
    F--H) sagittal (upper) and coronal (lower) views of the simulated tracer concentrations overlayed on the T1-weighted MR image after 4 and 6h for the baseline (K), high PVS flow (L) and high PVS dispersion (M) models (concentration opacity linearly increasing with concentration value, pial surface outlined in pink, arachnoid membrane in cyan, and arteries in dark red).}
\end{figure}

@@

      Molecular influx, time-to-peak & $T_{\rm max}$ & $1$ & h & \fixme{$\ast$} \\

@@

All authors express their gratitude to Prof.~Jürgen Reichenbach for
generously sharing the multi-modal human MR data, and to Prof. Erlend
Hodneland for facilitating this data exchange. M.E.R. acknowledges the
support of the Foundation Kristian Gerhard Jebsen via the K.~G.~Jebsen
Center for Brain Fluid Research and the Research Council of Norway via
grant \#324239 (EMIx). Simulations were performed on the Experimental
Infrastructure for Exploration of Exascale Computing (eX3), which is
financially supported by the Research Council of Norway under contract
270053. M.K. acknowledges support from the RCN grant \#303362.

>>>>

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to
  human brain health, with alterations often linked to
  disease and functional impairment, and with distinct opportunities
  for personalized diagnostics and treatment. However, our
  understanding of these mechanisms and their interplay remains
  incomplete, in part due to the complexity of integrating insights
  across scales, between species and from different modalities. Here,
  we combine mixed-dimensional modelling, multi-modal magnetic 
  resonance images, and high performance computing to construct and 
  explore a high-fidelity in-silico model of human intracranial molecular enrichment. 
  This model predicts the temporo-spatial spreading of a solute
  within an image-derived geometric representation of the
  subarachnoid space, ventricular system and brain parenchyma,
  including networks of surface perivascular spaces
  (PVSs). Our findings highlight the significant impact of
  cerebrospinal fluid (CSF) production and intracranial pulsatility on
  molecular enrichment following intrathecal tracer injection. We
  demonstrate that low-frequency vasomotion induces moderate CSF flow
  in surface PVS networks that substantially enhances tracer
  enrichment, and that impaired enrichment is a direct
  consequence of enlarged PVSs. This openly available technology
  platform thus provides an opportunity for integrating separate
  observations on diffusion in neuropil, vascular dynamics,
  intracranial pulsatility, CSF production, and efflux, and for
  exploring drug delivery and clearance in the human brain.
\end{abstract}
\begin{document}

@@

Our periarterial network extends from the internal carotid arteries
and basilar artery through up to 18 bifurcations to reach upstream
network ends located within the SAS or up to 6 mm inside the
parenchyma (\Cref{fig:results1}A--C, \Cref{fig:pvs_a}A) and includes
major surface arteries of radius 0.5--1.4 mm. Imposing the pressure
field induced by CSF production at the periarterial network ends
induces slow steady CSF flow of variable direction in these PVSs,
with an average velocity of 0.08 \textmu m/s (antegrade) and a maximum
velocity of 0.55 \textmu m/s (\Cref{fig:pvs_a}B). On the other hand,
traveling waves of arterial wall motion (\Cref{fig:pvs_a}C), such as the pulse wave or other vasomotion~\cite{vanveluw2020vasomotion,
  munting2023spontaneous, bojarskaite2023sleep, broggini2024long,hauglund2025norepinephrine},
also induce net directional flow in the
PVS~\cite{kedarasetti2020arterial, kedarasetti2020functional,
  coenen2021lubrication, gjerde2023directional, nozaleda2024arterial}
-- of magnitude and direction depending on the amplitude, frequency,
and length of the waves and the characteristics of the perivascular
network~\cite{gjerde2023directional}. Applying a
semi-analytic model of the net flow induced by peristalsis in
perivascular networks~\cite{gjerde2023directional} (see Methods), we
estimate that the cardiac pulse wave alone, traveling at a frequency
of 1 Hz with a wavelength of 2.0 m and a 1\% wall
displacement~\cite{jung2021novel}, will induce mainly antegrade PVS
flow with an average net velocity of 0.92 \textmu m/s while reaching up
to 7.31 \textmu m/s near larger, ventral vessels such as the MCA
(\Cref{fig:pvs_a}D). The same theory predicts that strong ultraslow
vasomotions, if traveling antegrade at 0.1Hz with a wavelength of 0.02
m and a 10\% displacement~\cite{broggini2024long}, will induce both
retrograde and antegrade net PVS flow with an average velocity of
13.05 \textmu m/s and maximum velocity 54.44 \textmu m/s
(\Cref{fig:pvs_a}E).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.99\textwidth]{figures/figure3b.png}
    \caption{\textbf{Perivascular flow shapes and accelerates molecular enrichment.}
    A) Comparison of mean tracer concentration in the PVS and over the outer PVS surface at key locations of the arterial tree, for the baseline (lower) and high PVS flow (upper) models, with labels: ``peak'': time-to-peak (h), ``$\Delta t$'': time difference between peaks (h), and ``FTA'': first-time tracer arrival (h).
    B) first-time arrival for the main trunks of the periarterial network over the distance from the closest network root node (BA, ICA-R or ICA-L for the baseline, the high PVS flow and the high PVS dispersion models);}
    \label{fig:pvs_b}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
    C, D) 2D slices showing zoom-in on the region surrounding the MCA-R and MCA2-R for the baseline (upper) and high PVS flow (lower) models. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF));  
    E) Total amount of tracer in the periarterial spaces as a function of the distance to the nearest network root over time for the baseline (left) and the high PVS flow (right) models;
    F--H) sagittal (upper) and coronal (lower) views of the simulated tracer concentrations overlayed on the T1-weighted MR image after 4 and 6h for the baseline (K), high PVS flow (L) and high PVS dispersion (M) models (concentration opacity linearly increasing with concentration value, pial surface outlined in pink, arachnoid membrane in cyan, and arteries in dark red). An interactive visualization of the baseline and the high PVS flow model can be found \href{https://baseline-pvs-flow.surge.sh/}{here}, and \href{https://high-pvs-flow.surge.sh/}{here}.}
\end{figure}

@@

      Molecular influx, time-to-peak & $T_{\rm max}$ & $1$ & h & $\ast$ \\

@@

All authors express their gratitude to Prof.~Jürgen Reichenbach for
generously sharing the multi-modal human MR data, and to Prof. Erlend
Hodneland for facilitating this data exchange. M.E.R. acknowledges the
support of the Foundation Kristian Gerhard Jebsen via the K.~G.~Jebsen
Center for Brain Fluid Research and the Research Council of Norway via
grant \#324239 (EMIx). Simulations were performed on the Experimental
Infrastructure for Exploration of Exascale Computing (eX3), which is
financially supported by the Research Council of Norway under contract
270053. 
M.C. acknowledges support from the RCN grant \#324239 (EMIx).
M.K. acknowledges support from the RCN grant \#303362.
The figures 1A, 1D, 2C, 3C, 5A and 6A were created in Biorender.
